---
title: "Publications"
editor_options: 
  chunk_output_type: console
---

## Publication Metrics

Information extracted from Google Scholar Profile

```{r ,echo=F, warning=F}

library(scholar)

id="MTXcoJYAAAAJ"

l=get_profile(id)

ct=get_citation_history(id)
```

`r paste0("Google Scholar Citations = ",l$total_cites)` <br> `r paste0("H-Index = ",l$h_index)` <br> `r paste0("i10-Index = ",l$i10_index)` <br>

*Information and definitions of various [indices](https://en.wikipedia.org/wiki/Author-level_metrics)*

## Books

## Peer Review Literature

```{r ,echo=F, warning=F}
## https://ropensci.org/blog/2020/05/07/rmd-citations/
library(RefManageR)
peer = RefManageR::ReadBib("./Bib/20231216_peer.bib",check=FALSE)

```

Carey, J., K. Jankowski, P. Julian, L. Sethna, P. Thomas, and J. J. Rohweder. 2019. Exploring Silica Stoichiometry on a Large Floodplain Riverscape. Frontiers in Ecology and Evolution 7: 346. Husk, B., P. Julian, N. Tromas, D. Phan, K. Painter, H. M. Baulch, and S. Sauvé. 2023. Improving Water Quality in a Hypereutrophic Lake and Tributary Through Agricultural Nutrient Mitigation: A Multi-Year Monitoring Analysis.doi:10.2139/ssrn.4572355 Jankowski, K. J., K. Johnson, L. Sethna, and others. 2023. Long-Term Changes in Concentration and Yield of Riverine Dissolved Silicon From the Poles to the Tropics. Global Biogeochemical Cycles 37: e2022GB007678. doi:10.1029/2022GB007678 Julian, P. 2011. Home range dynamics of female Florida panthers in reponse to kitten production. Florida Scientist 74. Julian, P. 2013a. Comment on Spatial and temporal phosphorus distribution changes in a large wetland ecosystem by X. Zapata-Rios et al.: Commentary. Water Resources Research 49: 2312--2313. doi:10.1002/wrcr.20162 Julian, P. 2013b. Mercury Bio-concentration Factor in Mosquito Fish (Gambusia spp.) in the Florida Everglades. Bulletin of Environmental Contamination and Toxicology 90: 329--332. doi:10.1007/s00128-012-0939-6 Julian, P. 2013c. Mercury hotspot identification in Water Conservation Area 3, Florida, USA. Annals of GIS 19: 79--88. doi:10.1080/19475683.2013.782469 Julian, P. 2014. Reply to Mercury Bioaccumulation and Bioaccumulation Factors for Everglades Mosquitofish as Related to Sulfate: A Re-Analysis of Julian II (2013). Bulletin of Environmental Contamination and Toxicology 93: 517--521. doi:10.1007/s00128-014-1389-0 Julian, P. 2015. South Florida Coastal Sediment Ecological Risk Assessment. Bulletin of environmental contamination and toxicology 95: 188--193. Julian, P. 2016. Commentary on Mitsch et al., 2015, Protecting the Florida Everglades wetlands with wetlands: Can stormwater phosphorus be reduced to oligotrophic conditions? Ecological Engineering 108: 333--337. Julian, P. 2017a. Assessment of Upper Taylor Slough water quality and implications for ecosystem management in Everglades National Park. Wetlands Ecology and Management 25: 191--209. doi:10.1007/s11273-016-9509-8 Julian, P. 2017b. Letter to the Editor regarding Surratt D, Shindle D, Yongshan W, et al. Letter to the Editor regarding: Julian P, 2017. Assessment of Upper Taylor Slough water quality and implications for ecosystem management in Everglades National Park. Wetlands Ecol Manage 1--3. doi:10.1007/s11273-017-9571-x Julian, P. 2019. Spatial Ecology and Conservation Modeling. Applications with R. Robert Fletcher, Marie-Josee Fortin: Book Review. Austral Ecology. doi:10.1111/aec.12791 Julian, P. 2020. Getting the science right to protect and restore our environment. A critique of Lapointe et al. (2019) Nitrogen enrichment, altered stoichiometry, and coral reef decline at Looe Key, Florida Keys, USA: a 3-decade study. Mar Biol 167: 68. doi:10.1007/s00227-020-3667-1 Julian, P., R. Chambers, and T. Russell. 2017a. Iron and Pyritization in Wetland Soils of the Florida Coastal Everglades. Estuaries and Coasts 40: 822--831. doi:10.1007/s12237-016-0180-3 Julian, P., and M. W. Cunningham. 2013. Total mercury concentration in Florida black bear (Ursus americanus floridanus). Florida Scientist 76. Julian, P., E. M. Everham III, and M. B. Main. 2012. Influence of a Large-scale Removal of an Invasive Plant (Melaleuca quinquenervia) on Home-range Size and Habitat Selection by Female Florida Panthers (Puma concolor coryi) within Big Cypress National Preserve, Florida. Southeastern Naturalist 11: 337--348. Julian, P., J. Fourqurean, S. Davis, and others. Submitted. Long-term spatiotemporal patterns and trends in water quality reveal a coastal continuum of disturbance legacies. Limnology and Oceanography. doi:https://doi.org/10.21203/rs.3.rs-1753636/v1 Julian, P., S. Gerber, R. K. Bhomia, J. King, T. Z. Osborne, and A. L. Wright. 2020. Understanding stoichiometric mechanisms of nutrient retention in wetland macrophytes: stoichiometric homeostasis along a nutrient gradient in a subtropical wetland. Oecologia. doi:10.1007/s00442-020-04722-9 Julian, P., S. Gerber, R. K. Bhomia, J. King, T. Z. Osborne, A. L. Wright, M. Powers, and J. Dombrowski. 2019. Evaluation of nutrient stoichiometric relationships among ecosystem compartments of a subtropical treatment wetland. Do we have "Redfield wetlands"? Ecol Process 8: 20. doi:10.1186/s13717-019-0172-x Julian, P., S. Gerber, A. L. Wright, B. Gu, and T. Z. Osborne. 2017b. Carbon pool trends and dynamics within a subtropical peatland during long-term restoration. Ecol Process 6: 43--57. doi:10.1186/s13717-017-0110-8 Julian, P., and B. Gu. 2015. Mercury accumulation in largemouth bass (Micropterus salmoides Lacépède) within marsh ecosystems of the Florida Everglades, USA. Ecotoxicology 24: 202--214. doi:10.1007/s10646-014-1373-9 Julian, P., B. Gu, and G. Redfield. 2015. Comment on and Reinterpretation of Gabriel et al. (2014) Fish Mercury and Surface Water Sulfate Relationships in the Everglades Protection Area. Environmental Management 55: 1--5. doi:10.1007/s00267-014-0377-9 Julian, P., B. Gu, and A. L. Wright. 2016a. Mercury Stoichiometric Relationships in a Subtropical Peatland. Water Air Soil Pollut 227: 472. doi:10.1007/s11270-016-3180-9 Julian, P., and T. Z. Osborne. 2018. From lake to estuary, the tale of two waters: A study of aquatic continuum biogeochemistry. Environment Monitoring and Assessment 190: 1--24. doi:https://doi.org/10.1007/s10661-017-6455-8 Julian, P., T. Z. Osborne, R. K. Bhomia, and O. Villapando. 2021. Knowing your limits: evaluating aquatic metabolism in a subtropical treatment wetland. Hydrobiologia. doi:10.1007/s10750-021-04617-7 Julian, P., T. Z. Osborne, and R. Ellis. 2023a. Evaluation of Biogeochemical Changes in Channelized and Restored Portions of a Subtropical Floodplain. Hydrobiology 2: 1--18. doi:10.3390/hydrobiology2010001 Julian, P., and L. Reidenbach. Submitted. Upstream water management and its role in estuary health, evaluation of freshwater management and subtropical estuary function. Water Resources Management. Julian, P., T. Schafer, M. J. Cohen, P. Jones, and T. Z. Osborne. 2023b. Changes in the spatial distribution of total phosphorus in sediment and water column of a shallow subtropical lake. Lake and Reservoir Management 39: 213--230. Julian, P., M. Thompson, and E. C. Milbrandt. 2024. Dark waters: Evaluating seagrass community response to optical water quality and freshwater discharges in a highly managed subtropical estuary. Regional Studies in Marine Science 69: 103302. doi:10.1016/j.rsma.2023.103302 Julian, P., and Z. Welch. 2022. Understanding the ups and downs, application of hydrologic restoration measures for a large Subtropical Lake. Lake and Reservoir Management 38: 304--317. doi:https://doi.org/10.21203/rs.3.rs-1739423/v2 Julian, P., A. L. Wright, and T. Z. Osborne. 2016b. Iron and sulfur porewater and surface water biogeochemical interactions in subtropical peatlands. Soil Science Society of America Journal 80: 794--802. Kominoski, J. S., E. E. Gaiser, E. Castaneda‐Moya, and others. 2020. Disturbance legacies increase and synchronize nutrient concentrations and bacterial productivity in coastal ecosystems. Ecology 101. doi:https://doi.org/10.1002/ecy.2988 Marazzi, L., C. M. Finlayson, P. A. Gell, P. Julian, J. S. Kominoski, and E. E. Gaiser. 2018. Balancing wetland restoration benefits to people and nature. Solutions Journal 9. Schafer, T. B., P. Julian, O. Villapando, and T. Z. Osborne. 2023. Abiotic mineralization of dissolved organic phosphorus for improved nutrient retention in a large-scale treatment wetland system. Ecological Engineering 195: 107078. doi:10.1016/j.ecoleng.2023.107078 Schafer, T., N. Ward, P. Julian, K. R. Reddy, and T. Z. Osborne. 2020. Impacts of Hurricane Disturbance on Water Quality across the Aquatic Continuum of a Blackwater River to Estuary Complex. Journal of Marine Science and Engineering 8: 412. doi:10.3390/jmse8060412 Smith, M. C., P. Julian, D. DeAngelis, and B. Zhang. 2023. Ecological benefits of integrative weed management of Melaleuca quinquenervia in Big Cypress National Preserve. BioControl. doi:10.1007/s10526-023-10229-y

## Technical Literature
