---
title: "Rejection"


---

One thing is certain that rejection is often called the constant companion in science. The collective we often focus on the rejection and don't celebrate the successes sufficently. If you search for any statistics regarding applying for grants, post-docs, jobs, etc. you will find some staggering statistics of number of applications, percent funded, etc. The old expression "Failure is not an option..." is more like failure is constantly an option so much so that back in 2016 [Dr Johannes Haushofer](https://www.princeton.edu/~joha/) became widely know for posting online his [CV of Failures](https://www.princeton.edu/~joha/Johannes_Haushofer_CV_of_Failures.pdf) others have follow suite. An [article](https://www.theguardian.com/education/2016/apr/30/cv-of-failures-princeton-professor-publishes-resume-of-his-career-lows) recounts that Haushofer, wrote the document to *"give some perspective"*. Most times we fail but it is hidden away out-of-sigh so when we do succeed it gives an unbalance representation of the day-to-day work. Another quote from the article helps puts things into context. 

>“As a result, they are more likely to attribute their own failures to themselves, rather than the fact that the world is stochastic, applications are crapshoots, and selection committees and referees have bad days.”

Motivated by Dr Haushofer, [Dr Melanie Stefan](http://melaniestefan.net/Mela_24Nov2017.pdf), [Dr Auriel Fournier](http://aurielfournier.github.io/rejection/) and others I decided to begin documenting my failures as well. This page is organized into three section including publications (*in-progress*), jobs, applications to workshops and grants/funding. I am sure I missed some data in reconstruction efforts but this accounts for the major applications and funding oppurtunities data from 2014/2015 to current.

<br>

## Jobs 

`r source("D:/commonlyUsedFunctions.R")`
I have been lucky that in the past ~ `r round(as.numeric(difftime(date.fun(Sys.Date()),date.fun("2011-08-01"),units="days"))/365,0)` years that I have held a very good position that has afforded me the chance to pursue multiple career development opportunities including my Professional Wetland Scientist (PWS) and doctorate. Despite these benefits and the freedom of my current position occasionally I search the job-boards and apply to positions that seem to fit my career trajectory. THIS IS PERFECTLY HEALTHY. However, some employers see this as bad because they don't want people looking to exit, using this job as a stepping stone, not be 100% engaged, etc. I see it as opportunities to advance my career, stay fresh with interviewing skills (not just for jobs) and potentially grow my network.  


```{r,echo=F}
#source("D:/commonlyUsedFunctions.R")
library(openxlsx)
library(plyr)

jobs=read.xlsx("D:/UF/RejectionTracker.xlsx",sheet=1)
jobs$Date=convertToDate(jobs$Date)
jobs$yr=format(jobs$Date,"%Y")

jobs.sum=ddply(jobs,"yr",summarise,N.total=length(Type),N.Inter=sum(Interviewed=="Y",na.rm=T),N.offer=sum(Offer=="Y",na.rm=T),N.dec=sum(Declined=="Y",na.rm=T))


```


```{r jobplot,echo=F,fig.width=5.5,fig.height=4,fig.align="center",fig.cap="Total number of job applications, interviews and offer during my job searches. * = Current Year"}
par(family="serif",mar=c(2,2,1,1),oma=c(1.5,2,0.5,1));

ylim.val=c(0,10);by.y=2;ymaj=seq(ylim.val[1],ylim.val[2],by.y);ymin=seq(ylim.val[1],ylim.val[2],by.y/2)
xlim.val=c(2015,2019,1);xlab=seq(xlim.val[1],xlim.val[2],1)
x=barplot(t(jobs.sum[,2:4]),beside=T,ylim=ylim.val,yaxt="n",col=grey.colors(3))
#axis_fun(1,x[2,]+(x[3,]-x[2,])/2,x[2,]+(x[3,]-x[2,])/2,seq(2015,2019,1))
axis_fun(1,x[2,],x[2,],paste0(xlab,c(rep(" ",length(xlab)-1),"*")))
axis_fun(2,ymaj,ymin,ymaj);box(lwd=1)
legend("topleft",legend=c("Total Applications","Total Interviews","Total Offers"),pch=22,pt.bg=grey.colors(3),pt.cex=1.5,ncol=1,cex=1,bty="n",y.intersp=1,x.intersp=0.75,xpd=NA,xjust=0.5)
mtext(side=1,line=2,"Year")
mtext(side=2,line=2,"Count")
```

<br>

## Workshops/Short-Courses

Professional development workshops, working groups and/or short courses sometimes accompany funding for travel expenses. Therefore, you are evaluated based on productivity, career, etc. like a job application.

```{r,echo=F}
work=read.xlsx("D:/UF/RejectionTracker.xlsx",sheet=2)
work$Date=convertToDate(work$Date)
work$yr=format(work$Date,"%Y")

work.sum=ddply(work,"yr",summarise,N.total=length(Type),N.offer=sum(Offer=="Y",na.rm=T),N.tbd=sum(is.na(Offer)))
work.sum=rbind(work.sum,data.frame(yr=2015:2017,N.total=0,N.offer=0,N.tbd=0))
work.sum=work.sum[order(work.sum$yr),]
```

```{r workplot,echo=F,fig.width=5.5,fig.height=4,fig.align="center",fig.cap="Total number of workshop applications, successful application and applications to-be-determined. * = Current Year"}
par(family="serif",mar=c(2,2,1,1),oma=c(1.5,2,0.5,1));

ylim.val=c(0,5);by.y=1;ymaj=seq(ylim.val[1],ylim.val[2],by.y);ymin=seq(ylim.val[1],ylim.val[2],by.y/2)
xlim.val=c(2015,2019,1);xlab=seq(xlim.val[1],xlim.val[2],1)
x=barplot(t(work.sum[,2:4]),beside=T,ylim=ylim.val,yaxt="n",col=grey.colors(3),xaxt="n")
#axis_fun(1,x[2,]+(x[3,]-x[2,])/2,x[2,]+(x[3,]-x[2,])/2,seq(2015,2019,1))
axis_fun(1,x[2,],x[2,],paste0(xlab,c(rep(" ",length(xlab)-1),"*")))
axis_fun(2,ymaj,ymin,ymaj);box(lwd=1)
legend("topleft",legend=c("Total Applications","Total Successful Applications","Total to-be-determined"),pch=22,pt.bg=grey.colors(3),pt.cex=1.5,ncol=1,cex=1,bty="n",y.intersp=1,x.intersp=0.75,xpd=NA,xjust=0.5)
mtext(side=1,line=2,"Year")
mtext(side=2,line=2,"Count")
```

<br>

## Funding

The next big category is funding including grants, scholarships, fellowships, etc.  

```{r,echo=F}
fund=read.xlsx("D:/UF/RejectionTracker.xlsx",sheet=3)
fund$Date=convertToDate(fund$Date)
fund$yr=format(fund$Date,"%Y")

fund.sum=ddply(fund,"yr",summarise,N.total=length(Type),N.offer=sum(Offer=="Y",na.rm=T),N.tbd=sum(is.na(Offer)))
fund.sum=rbind(fund.sum,data.frame(yr=2019,N.total=0,N.offer=0,N.tbd=0))
```

```{r fundplot,echo=F,fig.width=5.5,fig.height=4,fig.align="center",fig.cap="Total number of funding applications, successful funding and applications to-be-determined. * = Current Year"}
par(family="serif",mar=c(2,2,1,1),oma=c(1.5,2,0.5,1));

ylim.val=c(0,5);by.y=1;ymaj=seq(ylim.val[1],ylim.val[2],by.y);ymin=seq(ylim.val[1],ylim.val[2],by.y/2)
xlim.val=c(2014,2019,1);xlab=seq(xlim.val[1],xlim.val[2],1)
x=barplot(t(fund.sum[,2:4]),beside=T,ylim=ylim.val,yaxt="n",col=grey.colors(3),xaxt="n")
#axis_fun(1,x[2,]+(x[3,]-x[2,])/2,x[2,]+(x[3,]-x[2,])/2,seq(2015,2019,1))
axis_fun(1,x[2,],x[2,],paste0(xlab,c(rep(" ",length(xlab)-1),"*")))
axis_fun(2,ymaj,ymin,ymaj);box(lwd=1)
legend("topleft",legend=c("Total Applications","Total Successful Applications","Total to-be-determined"),pch=22,pt.bg=grey.colors(3),pt.cex=1.5,ncol=1,cex=1,bty="n",y.intersp=1,x.intersp=0.75,xpd=NA,xjust=0.5)
mtext(side=1,line=2,"Year")
mtext(side=2,line=2,"Count")
```